---
title: "KFS Parameter Initialization"
author: "Ewok"
date: "October 10, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Pseudocode

### Input

y - m x t, observations/measurements  

### Output

A - n x n, state transition matrix  
C - m x n, observation model matrix  
Q - n x n, environmental noise  
R - m x m, measurement noise  
x0 - n x 1, initial state estimate  
v0 - n x n, initial covariance estimate  

### Function Names

KFS_init - initializes the parameters for the Kalman Filter/Smoother

### Instructions

Read the paper found [here](http://numbat.ucsd.edu/~bob/estimation/Estimation_Fall_2010/Notes_and_docs_files/P61_Barb_TSP.pdf) for better understanding of importance of initial noise estimates.

We will initialize the state estimate to be the first measurement given by the input y, because this is simplest estimate. We will initialize the covariance estimate to the identity matrix, because v0 is less crucial to the outcome of the KFS than Q and R, and is typically an underestimate. Too small an estimate for Q leads to overconfidence in the predicted state, while too large an estimate leads to oversensitivity to noise. Q determines the amount of weight given to the Kalman gain. We will initialize Q to the identity matrix. The choice of R is less important, as it has more of an impact on the variation of the state estimates. We will initialize R to the identity matrix times the estimated mean squared error, which we define as the following: $I\times\frac{1}{n}\sum_{t=1}^{T-2}\left(y_{t+1}-\frac{y_t+y_{t+2}}{2}\right)^2$. Given no information about the system, we will assume the state variables are independent and linear, thus initializing A to the identity matrix. We will assume C is a diagonal matrix of 1s (not necessarily square).

## Simulation Description

## Evaluating Performance

## Algorithm Code

```{r kfs_init}
KFS_init <- function(y) {
  n = dim(y)[1]
  t = dim(y)[2]
  A <- diag(n)
  C <- diag(n)
  Q <- diag(n)
  sum = 0
  for(i in 1:t-2) {
    sum = sum + (y[i+1]-((y[i]+y[i+2])/2))^2
  }
  R <- I * (sum/n)
  x0 <- y[1]
  v0 <- diag(n)
  return(list(A=A, C=C, Q=Q, R=R, x0=x0, v0=v0))
}
```
